---
title: "GSA 2022 Abstract Code"
author: "Peter Sun"
date: "July 30, 2022"
output:
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
header-includes:
- \usepackage{enumitem}
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---

```{r setup, include=FALSE}
options(scipen=999)
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = "") # omit NAs in Kable
```

\newpage
# Load Packages and Data

```{r message=F, error=F, warning=F}
library(tidyverse)
library(haven)
library(sjlabelled)
library(ggpubr)
library(kableExtra)
library(scales)
library(choroplethr)
library(choroplethrMaps)
library(glue)
library(tigris)
data(state.regions)
library(janitor)
library(fips)
library(cowplot)
library(biscale)

# Avoid select clashes
select <- dplyr::select
recode <- dplyr::recode
summarize <- dplyr::summarize
```

\newpage
# Re-Read Project-Level Environmental Variables for HRS Dataset File Paths

```{r}
readRenviron(".Renviron")
```

\newpage
# Import Work, Age, Wave 14 (2018) Flag, and Nursing Home Status

```{r}
# Import "randhrs1992_2018v1.dta"
rand.long <- read_dta(Sys.getenv("HRS_LONG"), 
  col_select = c(hhid, pn, r14agey_e, inw14,
                 r14work, # Currently Working for Pay
                 s14hhidpn)) %>% # Spouse Identifier for Caregiving
  rename(worker = r14work) %>%
  haven::zap_formats() %>%
  sjlabelled::remove_all_labels() %>%
  as_tibble()

# Inspect worker
table(rand.long$worker, useNA = "always") # 6711 workers

# Import nursing home status from "trk2018tr_r.dta"
tracker <- read_dta(Sys.getenv("HRS_TRACKER_2018_20"),
  col_select = c(hhid, pn, qnurshm))
```

\newpage
# Import Geography Data

```{r}
# Import "HRSXREGION18.dta"
geo <- read_dta(Sys.getenv("HRS_REGION_2018_82"),
    col_select = c(hhid, pn, beale2013_18, region18)) %>%
  rename(rural = beale2013_18) %>%
  mutate(rural = recode(rural, 
    `1` = "Urban",
    `2` = "Urban",
    `3` = "Rural",
    .default = NA_character_)) %>%
  mutate(region = recode(region18, 
    `1` = "northeast", 
    `2` = "northeast",
    `3` = "midwest",
    `4` = "midwest",
    `5` = "south", 
    `6` = "south",
    `8` = "west",
    `9` = "west", 
    .default = NA_character_)) %>%
  mutate(division = recode(region18,
    `1` = "New England",
    `2` = "Middle Atlantic",
    `3` = "East North Central",
    `4` = "West North Central",
    `5` = "South Atlantic",
    `6` = "East South Central",
    `7` = "West South Central",
    `8` = "Mountain",
    `9` = "Pacific", .default = NA_character_)) %>%
  haven::zap_formats() %>%
  sjlabelled::remove_all_labels() %>%
  as_tibble()
```

\newpage
# Import Volunteering Data

```{r}
vol.18 <- read_dta(Sys.getenv("HRS_2018_FAT"),
    col_select = c(hhid, pn, qg086)) %>%
  rename(volunteer = qg086) %>%
  mutate(volunteer = recode(volunteer,
    `1` = 1,
    `5` = 0, 
    .default = NA_real_))
table(vol.18$volunteer, useNA = "always") # 5772 volunteers
```

\newpage
# Import Caregiving Data

```{r}
# Import 2018 RAND Fat File (File name: h04f1c.dta)
care18 <- read_dta(Sys.getenv("HRS_2018_FAT"),
  col_select = c(
    "hhidpn", "hhid", "pn",

    # adl helpers
    starts_with("qg033_"),

    # iadl helpers
    starts_with("qg055_"),

    # caregiving grandchildren
    "qe060",

    # caregiving parental personal
    "qf119",

    # caregiving parental errands
    "qf139"
  )
) %>%
  haven::zap_formats() %>%
  sjlabelled::remove_all_labels() %>%
  as_tibble()

# Identify participants who had a spouse/partner ADL or IADL helper
# QG033_x = ADL helpers
# QG055_x = IADL helpers
# TODO: those who did not report having ADL or IADL needs should be marked as
# not having an ADL/IADL caregiver
care18b <- care18 %>%
  # Temporarily set 2s to 1, non-2 numbers to 0, and keep NAs as NAs
  # Then count the number of 1s (which is equivalent to the number of 2s)
  mutate(spouse_helper_sum = rowSums(
    ifelse(select(., starts_with("qg033"), starts_with("qg055")) == 2, 1, 0),
    na.rm = T
  )) %>%
  
  # Because rowSums in the above syntax will count all NAs as 0, 
  # use !is.na(...) == 0 to set all
  # non-missing values to TRUE and missing values to FALSE
  # Then use rowSums to count the number of non-missing values
  # Because FALSE = missing, if all of the columns are FALSE/NA/Missing, 
  # then the sum will be 0
  # Finally, set spouse_helper_sum to NA_real_ if all of the columns are False/NA/Missing
  mutate(spouse_helper_sum = ifelse(
    rowSums(!is.na(select(., starts_with("qg033"), starts_with("qg055")))) == 0, 
    NA_real_, spouse_helper_sum
  )) %>%
  
  # spouse_helper_sum counts the number of 2s (spouse/partner) in QG033_x and QG055_x
  # has_spouse_helper coding:
  # 1 = if there is at least one 2s across all QG033_x and QG055_x variables
  # 0 = if there is at least one non-missing variable and no 2s
  # NA = all QG033_x and QG055_x variables are NAs
  mutate(has_spouse_helper = ifelse(spouse_helper_sum >= 1, 1, 
                             ifelse(is.na(spouse_helper_sum), NA_real_, 0)))
table(care18b$has_spouse_helper, useNA = "always")

# Extract participants who have a spousal ADL/IADL caregiver
# Merge their spouse PN
# Then create a dataset with hhid and pn of spouse and an indication of
# whether or not they are a spousal caregiver
spousal_caregivers.18 <- care18b %>%
  filter(has_spouse_helper == 1) %>%
  select(hhid, pn, has_spouse_helper) %>%
  
  # Left join with respondents' spouse/partner's hhidpn in wave 14 (2018)
  # "If there is no spouse in a given wave, SwHHIDPN is set to zero. If SwHHIDPN is 
  # unknown, and the marital status in a particular wave
  # is either missing (.M) or married, SwHHIDPN is set to a special missing code of .M."
  left_join(rand.long %>% select(hhid, pn, s14hhidpn) %>% 
              filter(!is.na(s14hhidpn), s14hhidpn != 0, !is.na(hhid), !is.na(pn)),
    by = c("hhid", "pn")
  ) %>%
  select(hhidpn = s14hhidpn, caregiver_spousal = has_spouse_helper)
spousal_caregivers.18
table(is.na(spousal_caregivers.18$hhidpn), useNA = "always") # 27 spouses not in dataset

# Merge the spousal_caregivers data back to the dataset
care18c <- care18b %>%
  left_join(spousal_caregivers.18, by = c("hhidpn"))

# Format parental/grandchildren caregivers
# qe060 = grandchildren caregiver
# qf119 = parental caregiver (personal needs) (1 = yes, 5 = no, 8 = DK, 9 = RF)
# qf139 = parental caregiver (errands)
care18d <- care18c %>%
  mutate(across(
    .cols = c(qe060, qf119, qf139),
    ~ recode(.,
      `1` = 1, `5` = 0, `8` = NA_real_, `9` = NA_real_,
      .default = NA_real_
    )
  )) %>%
  rename(
    caregiver_grandchildren = qe060,
    caregiver_parental_personal = qf119,
    caregiver_parental_errands = qf139
  ) %>%
  mutate(
    caregiver_parental =
      ifelse(caregiver_parental_personal == 1 | caregiver_parental_errands == 1, 1, 
      ifelse(is.na(caregiver_parental_personal) & is.na(caregiver_parental_errands), 
             NA_real_, 0)))
table(care18d$caregiver_parental, useNA = "always")

# Caregiver Coding
# 1 = If at least one type of caregiver
# NA = Two or more NAs
# 0 = All other cases (at least two zeroes)
care18e <- care18d %>%
  select(hhid, pn,
    cs = caregiver_spousal,
    cp = caregiver_parental,
    cg = caregiver_grandchildren
  ) %>%
  mutate(
    Caregiver_Sum = rowSums(select(., cs:cg), na.rm = T),
    # Set caregiver to NA if two or more NAs
    Caregiver_NACount = rowSums(is.na(select(., cs:cg))),
    Caregiver_Sum = ifelse(Caregiver_NACount >= 2, NA_real_, Caregiver_Sum),
    caregiver = ifelse(Caregiver_Sum >= 1, 1, 
                ifelse(is.na(Caregiver_Sum), NA_real_, 0))
  ) %>%
  select(hhid, pn, caregiver)
table(care18e$caregiver, useNA = "always") # 2640 caregivers
```

\newpage
# Merge Datasets and Create Multiple Productive Activities

```{r}
df <- rand.long %>%
  left_join(tracker, by = c("hhid", "pn")) %>%
  filter(inw14 == 1) %>% # in wave 14
  filter(qnurshm %in% c(5, 6, 7)) %>% # community-dwelling
  filter(r14agey_e >= 65) %>% # age 65+
  left_join(geo, by = c("hhid", "pn")) %>%
  left_join(vol.18, by = c("hhid", "pn")) %>%
  left_join(care18e, by = c("hhid", "pn")) %>%
  
  # Count 0s and 1s
  mutate(multi_zeroes = rowSums(select(., volunteer, caregiver, worker) == 0, 
                                na.rm = T)) %>%
  mutate(multi_ones = rowSums(select(., volunteer, caregiver, worker) == 1, 
                              na.rm = T)) %>%
  
  # Set multi to 0 if at least one 0, otherwise NA
  mutate(multi = ifelse(multi_zeroes >= 1, 0, NA_real_)) %>%
  
  # Set multi to 1 if at least one productive activity
  mutate(multi = ifelse(multi_ones >= 1, 1, multi))
  # rowwise() %>%
  # mutate(multi_sum = sum(c_across(c(volunteer, caregiver, worker)), na.rm = T)) %>%
  # ungroup()

table(df$volunteer, useNA = "always")
table(df$caregiver, useNA = "always") # likely an underestimation of non-caregivers
table(df$worker, useNA = "always")  
table(df$multi, useNA = "always")
#df %>% select(volunteer, caregiver, worker, multi_zeroes, multi_ones, multi) %>% view()
```

## Study Sample Size

```{r}
study_n <- nrow(df)
study_n # 2018 HRS sample of age 65+ community-dwelling individuals
```

\newpage
# Helper Functions

```{r}
# Function for contingency table
# The denominator is currently assumed to be the sum of 0s and NAs, that is,
# the total population in each region/division. This is a 
# limitation in this current study, because of the possibility of non-response bias.
get_kab <- function(data, geo, iv) {
  data %>%
    count({{ geo }}, rural, {{ iv }}) %>%
    group_by({{ geo }}, rural) %>%
    mutate(denom = sum(n),
           pct = n / denom) %>%
    ungroup() %>%
    filter({{ iv }} == 1) %>%
    filter(!is.na({{ geo }}), !is.na(rural), !is.na({{ iv }})) %>%
    select(-{{ iv }})
}

# Function for two proportion z-test
get_prop <- function(data, geo, iv) {
  data %>%
    count({{ geo }}, rural, {{ iv }}) %>%
    group_by({{ geo }}, rural) %>%
    mutate(sum = sum(n)) %>%
    ungroup() %>%
    filter({{ iv }} == 1) %>%
    filter(!is.na({{ geo }}), !is.na(rural), !is.na({{ iv }})) %>%
    pivot_wider(names_from = "rural", values_from = n:sum) %>%
    rowwise() %>%
    mutate(p = prop.test(x = c(n_Rural, n_Urban), 
                         n = c(sum_Rural, sum_Urban))$p.value,
           rural_prop = prop.test(x = c(n_Rural, n_Urban), 
                                  n = c(sum_Rural, sum_Urban))$estimate[1],
           urban_prop = prop.test(x = c(n_Rural, n_Urban), 
                                  n = c(sum_Rural, sum_Urban))$estimate[2]) %>%
    mutate(rural_prop = percent(rural_prop, accuracy = .1),
           urban_prop = percent(urban_prop, accuracy = .1)) %>%
    kbl(booktabs = T, linesep = "", digits = 3) %>%
    kable_styling(position = "center") %>%
    kable_styling(latex_options = c("striped", "hold_position")) 
}
```

\newpage
# Results

## Census Region Statistics

```{r}
r1 <- get_kab(df, region, worker) %>% rename(worker = pct)
r2 <- get_kab(df, region, volunteer) %>% rename(volunteer = pct)
r3 <- get_kab(df, region, caregiver) %>% rename(caregiver = pct)
r4 <- get_kab(df, region, multi) %>% rename(multiple = pct)
r1 %>%
  left_join(r2, by = c("region", "rural")) %>%
  left_join(r3, by = c("region", "rural")) %>%
  left_join(r4, by = c("region", "rural")) %>%
  select(-starts_with("n"), -starts_with("denom")) %>%
  mutate(across(where(is.numeric), scales::percent, 0.1)) %>%
  kbl(booktabs = T, linesep = "") %>%
  kable_styling(position = "center") %>%
  kable_styling(latex_options = c("striped", "hold_position"))  
```

\newpage
## Census Division Statistics

```{r}
d1 <- get_kab(df, division, worker) %>% rename(worker = pct)
d2 <- get_kab(df, division, volunteer) %>% rename(volunteer = pct)
d3 <- get_kab(df, division, caregiver) %>% rename(caregiver = pct)
d4 <- get_kab(df, division, multi) %>% rename(multiple = pct)
d1 %>%
  left_join(d2, by = c("division", "rural")) %>%
  left_join(d3, by = c("division", "rural")) %>%
  left_join(d4, by = c("division", "rural")) %>%
  select(-starts_with("n"), -starts_with("denom")) %>%
  mutate(across(where(is.numeric), scales::percent, 0.1)) %>%
  kbl(booktabs = T, linesep = "") %>%
  kable_styling(position = "center") %>%
  kable_styling(latex_options = c("striped", "hold_position"))  
```

\newpage
## Rural-Urban Comparisons (Two-Proportion Z Tests)

### Region: Worker, Volunteer, Caregiver, and Multiple

```{r}
get_prop(df, region, worker)
get_prop(df, region, volunteer)
get_prop(df, region, caregiver)
get_prop(df, region, multi)
```

\blandscape

\newpage
### Division: Worker and Volunteer

```{r}
get_prop(df, division, worker)
get_prop(df, division, volunteer)
```

\newpage
### Division: Caregiver and Multiple

```{r error=F, warning=F, message=F}
get_prop(df, division, caregiver)
get_prop(df, division, multi)
```

\elandscape

\newpage
# Plots

```{r}
# Obtain a list of states and FIPS codes
fp <- tidycensus::fips_codes %>%
  as_tibble() %>%
  group_by(state) %>%
  slice(1) %>%
  ungroup() %>%
  select(state_code = state, GEOID = state_code) %>%
  arrange(GEOID)

# Join FIPS codes with Census divisions
rd <- read_csv("private/poster/us_census_regions_and_divisions.csv") %>%
  janitor::clean_names() %>%
  left_join(fp, by = "state_code") %>%
  arrange(GEOID) %>%
  select(GEOID, division)

# Get list of states per division
read_csv("private/poster/us_census_regions_and_divisions.csv") %>%
  janitor::clean_names() %>%
  left_join(fp, by = "state_code") %>%
  group_by(division) %>%
  summarize(states = paste0(state_code, collapse = ", "))

# Get proportions of older adults with at least one productive activity by division
multi <- get_kab(df, division, multi) %>%
  select(-n, -denom) %>%
  pivot_wider(names_from = "rural", values_from = "pct")
multi

# Join data with FIPS/divisions
rd_multi <- rd %>%
  left_join(multi, by = "division")

# Download a states map (filter out Puerto Rico)
# Transform to USA Contiguous Albers Equal Area Conic ('ESRI:102003')
states_tmp <- tigris::states(class = "sf", resolution = "20m", cb = T) %>%
  filter(GEOID != "72") %>% # filter out Puerto Rico (PR)
  shift_geometry()

# Merge sf map with data
states <- states_tmp %>%
  left_join(rd_multi, by = "GEOID")

# Function to produce a bivariate choropleth map
gen_map <- function(mypal) {
  mydim <- 3
  mystyle <- "quantile"
  data <- bi_class(states, x = Rural, y = Urban, style = mystyle, dim = mydim)
  breaks <- bi_class_breaks(states, x = Rural, y = Urban, style = mystyle, dim = mydim)
  map <- ggplot() +
    geom_sf(
      data = data, mapping = aes(fill = bi_class), show.legend = FALSE,
      color = "gray65", size = 0.5
    ) +
    bi_scale_fill(pal = mypal, dim = mydim) +
    # geom_sf_label(data = data, aes(label = NAME)) +
    bi_theme(bg_color = "transparent") +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA),
      plot.title = element_text(color = "gray80", size = 30, face = "bold"),
      legend.background = element_rect(fill = "transparent", color = NA),
      legend.position = "bottom",
      plot.margin = margin(0, 200, 0, 0, "pt") # top, right, bottom, left
    )
  default_background_color <- "transparent"
  legend <- bi_legend(
    pal = mypal,
    dim = mydim,
    xlab = "Higher % PE in Rural",
    ylab = "Higher % PE in Urban",
    size = 12
  ) +
    theme(
      plot.background = element_rect(
        fill = default_background_color,
        color = NA
      ),
      panel.background = element_rect(
        fill = default_background_color,
        color = NA
      ),
      legend.background = element_rect(
        fill = default_background_color,
        color = NA
      ),
      text = element_text(color = "white")
    )
  finalPlot <- ggdraw() +
    draw_plot(map, 0, 0, 1, 1) +
    draw_plot(legend, 0.6, .1, 0.35, 0.35)
  finalPlot
}

# Produce map based on given bivariate color scheme
# pals <- c("Bluegill", "BlueGold", "BlueOr", "BlueYl", "Brown", "Brown2",
# "DkBlue", "DkBlue2", "DkCyan", "DkCyan2", "DkViolet", "DkViolet2", "GrPink",
# "GrPink2", "PinkGrn", "PurpleGrn", "PurpleOr")
pals <- "DkBlue"
for (p in pals) {
  # mypal <- "DkBlue2"
  mymap <- gen_map(p)
  print(mymap)
  map_filename <- glue("map_{p}.png")
  ggsave(map_filename,
    plot = mymap, width = 12, height = 7, dpi = 600,
    bg = "transparent"
  )
}
```